Liveness-Based Garbage Collection for Lazy Languages

### 2015-06-10 paper/paper.tex
Fixing the section on implementation schemes

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
%\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage [plain,noend,noline,boxed]{algorithm2e}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
%\usepackage{program}
\usepackage{algorithmic}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\cred}[1]{{\color{red}{#1}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
\newcommand{\stk}{\mbox{S}}       % stack
\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\begin{document}

% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Liveness-Based Garbage Collection for Lazy Languages}
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.
%\cmt{{
%\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors
section.
%
\author{
% 1st. author
\alignauthor K. Prasanna Kumar \\
       \affaddr{IIT Bombay,}\\
       \affaddr{Mumbai 400076, India}
       \email{prasanna@cse.iitb.ac.in}
% 2nd. author
\alignauthor Amey Karkare\\
       \affaddr{IIT Kanpur}\\
       \affaddr{Kanpur 208016, India}
       \email{karkare@cse.iitk.ac.in}
% 3rd. author
\alignauthor Amitabha Sanyal \\
       \affaddr{IIT Bombay}\\
       \affaddr{Mumbai 400076, India}\\
       \email{as@cse.iitb.ac.in}
}
}}
\authorinfo{Double Blind Review}{}{}
\maketitle

@@

While  the runtime  system  of most  functional  languages includes  a
garbage collector to efficiently  reclaim memory, empirical studies on
Scheme~\cite{karkare06effectiveness}   and,   more   importantly,   on
Haskell~\cite{rojemo96lag}
%and Java~\cite{shaham02estimating}
programs have shown that  garbage collectors leave uncollected a large
number of  memory objects that are  reachable but not  live (here {\em
  live} means the  object can potentially be used by  the program at a
later stage).  This results in unnecessary memory retention.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%% MOTIVATING EXAMPLE
\newcommand{\nilfigure}
{\scalebox{0.75}{
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,0)(1,1)
%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,
subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
\putnode{start}{origin}{0}{0}{}
\putnode{stop}{origin}{10}{10}{}
\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
\end{pspicture}
}}

\begin{figure}[t!]
  %%%%{\color{Myblue}
  \psset{unit=1mm}
  \begin{pspicture}(80,50)(0,-50)
    %\psframe(80,50)(0,-50)
    %  \begin{center}
    \begin{tabular}{@{}c@{}}
      \begin{boxedminipage}{.42\textwidth}
        {\sf
	  \renewcommand{\arraystretch}{1}{
	    \begin{uprogram}
              \UFL\
              \UNL{0} (\DEFINE\ (\length\  \pl)
	      \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
              (\PRIM\ 1\ (\CAR\  \pl))))
              \UNL{0}
	      \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
	      \UNL{1}  (\SIF~(\NULLQ \ \lista)
	      \listb
	      \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\  \lista) (\append\
              (\CDR\  \lista)
              \listb))))
              \UNL{0}
              \UNL{0} (\LET\ \px\
              $\leftarrow$\ (\CONS\ $5$
              %          \UNL{1}\hspace*{1.5cm}
              (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
	      \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
	      \UNL{2}
              (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
              \UNL{3} (\SIF~(\NULLQ~(\CAR~(\CDR~\pz)))~$0$~$\pi$:\
              (\length\ \pz))))))
	    \end{uprogram}
        }}
      \end{boxedminipage}
      \\
      (a) Example program. \\ \\
      \begin{minipage}{.45\textwidth}
        %\hspace*{1cm}
                {\epsfig{file=motiv-example.eps, height=4cm}}
      \end{minipage}
      \\ \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
          (b) Memory graph at $\pi$. Thick edges denote live
          links. Traversal \\{\white (b)} stops at edges
          marked $\times$ during garbage collection.
      \end{tabular}}
    \end{tabular}
  \end{pspicture}
 \vspace*{-3ex}
  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
\end{figure}

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The analysis  has two  parts.  First,  each  function is
summarized in a context-sensitive manner as a {\em demand transformer}
that transforms a {\em symbolic} demand  on its body to demands on its
arguments.  This summary  is then used to step  through function calls
during analysis.   Second, the concrete  demand on a function  body is
obtained through a 0-CFA-like conservative approximation that combines
the  demands on  all the  calls to  the function.   The result  of the
analysis  is the annotation  of each  program point  with finite-state
automata capturing  the liveness of variables at  the point. Depending
on the program point where  garbage collection has been triggered, the
garbage-collector consults  a set of automata  to curtail reachability
during marking.  This results in  an increase in the garbage reclaimed
and consequently in fewer garbage collections.

@@

The edges shown by thick arrows  are those which are live at $\pi$.  A
cell is marked and preserved  during garbage collection only if it is
reachable from the  root set through a path of  live edges.  All other
cells can be  reclaimed.  Thus if a garbage  collection takes place at
$\pi$ with the heap shown in Figure~\ref{fig:mot-example}(b), only the
cells referenced by $\pz$,  $(\CDR~ \pz)$ and (the cells constituting)
the closure  referenced by $(\CDR~(\CDR~\pz))$  need to be  marked and
preserved.  In  contrast, a  reachability-based  collector would  have
preserved all cells.

@@

The      $\mathit{ref}$     rule      for     \CONS,      shown     in
Figure~\ref{fig:live-judge},  requires   us  to  remove   the  leading
\acar\ and \acdr\  from the access paths in  $\sigma$.  Similarly, the
rules for  \CAR, \CDR, \PRIM, \NULLQ,  and \SIF\ require  us to return
$\emptyset$, if  $\sigma$ itself is $\emptyset$.  To  realize these rules
$\sigma$ needs to be known. This creates difficulties since we want to
solve the equations arising out of liveness symbolically.

The  solution  is  to   also  treat  the  operations  mentioned  above
symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
\bcdr\ are defined as:
\begin{align*}
 \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
 \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
\end{align*}
and \clazy\ as:
\begin{align*}
 \clazy\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
                       & \{\epsilon\} ~\mbox{otherwise}
\end{align*}
Note that \clazy\ is idempotent. 
We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\clazy\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,\mathit{ep} \mapsto  x.\clazy \sigma]
\end{align*}
The rules for \CDR, \PRIM\ and  \NULLQ\  are also modified in a manner
similar to \CAR.
The  defintions of \bcar\, \bcdr\ and \clazy\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \acdr\sigma_2',~\mbox{and}\\
&\sigma_1\clazy \sigma_2 \hookrightarrow \epsilon, \mbox{ if } \sigma \ne
  \emptyset 
\end{align*}
The reflexive  transitive closure of  $\hookrightarrow$ will
be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
following proposition relates  the original and the modified
liveness rules.

@@

\comment{A paragraph more to explain the illustration on the following lines:

%------------------------------------------------------------%
%% \subsubsection{Transforming NFAs to Accept Forward Paths:}
%% %\label{sec:NFA-elim}
%% The   strongly  regular   CFGs  obtained   after   the  Mohri-Nederhof
%% transformation are first converted into NFAs.  The algorithm described
%% in  Figure~\ref{algo:simplify-NFA} converts  an NFA  \ntbar\ to  a NFA
%% \nt\    such    that   $\lang{\ntbar}    \stackrel{*}{\hookrightarrow}
%% \lang{\nt}$, where {\nt} accepts forward paths only.  \comment{Rewrite
%%   the algorithm to handle \clazy\  and update the example automata} Thus
%% \nt\ can be used to check membership of forward paths.

%%\begin{figure*}[t]
%% \begin{boxedminipage}{.5\textwidth}
%% \begin{center}
%%   \raggedright  {\bf  Input:}  NFA \ntbar\  with  underlying
%%   alphabet $\lbrace \acar, \acdr, \bcar, \bcdr\rbrace$ \\
%% %
%%   \raggedright{\bf Output:} NFA  \nt\ with  underlying
%%   alphabet $\lbrace \acar, \acdr \rbrace$ such that
%%   $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$.
%%   \\

%%   \raggedright{\bf Steps:}
%%  \begin{algorithmic}
%%    \STATE $i \leftarrow 0$
%%    \STATE $\nt_0 \leftarrow$ Equivalent NFA of \ntbar\
%%    without $\epsilon$-moves \cite{hopcraft90toc}

%%    \REPEAT
%%    \STATE $\nt'_{i+1} \leftarrow \nt_i$
%%    \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
%%      edge from $q'$ with label $\bcar$ and
%%      outgoing edge to $q''$ with label $\acar$}
%%    \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
%%    label $\epsilon$.
%%    \COMMENT{bypass  $\bcar\acar$ using $\epsilon$}
%%    \ENDFOR

%%    \FORALL{states $q$ in $\nt_i$ such that $q$ has an incoming
%%      edge from $q'$ with label $\bcdr$ and outgoing edge to
%%      $q''$ with label $\acdr$}
%%    \STATE add an edge in $\nt'_{i+1}$ from $q'$ to $q''$ with
%%      label $\epsilon$.
%%    \COMMENT{bypass  $\bcdr\acdr$ using $\epsilon$}
%%    \ENDFOR

%%    \STATE   $\nt_{i+1}    \leftarrow$   Equivalent   NFA   of
%%    $\nt'_{i+1}$ without $\epsilon$-moves

%%    \STATE $i \leftarrow i+1$
%%    \UNTIL ($\nt_{i} = \nt_{i-1}$)

%%    \STATE $\nt \leftarrow \nt_i$
%%   \STATE delete all edges with label \bcar\ or \bcdr\ in \nt.
%%  \end{algorithmic}
%% \end{center}
%% \end{boxedminipage}
%% \begin{boxedminipage}{.5\textwidth}
%% The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
%% of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
%% is continued till  a fixed point is reached.  When  the fixed point is
%% reached,   the  resulting   NFA  contains   all   possible  reductions
%% corresponding to all the paths in the original NFA.  The proofs of the
%% termination and correctness of  th The algorithm repeatedly introduces
%% $\epsilon$  edges  to bypass  a  pair  of  consecutive edges  labelled
%% \bcar\acar\  or \bcdr\acdr.   The process  is continued  till  a fixed
%% point is reached.  When the  fixed point is reached, the resulting NFA
%% contains all possible reductions corresponding to all the paths in the
%% original NFA.  The proofs of the termination and correctness of th The
%% algorithm repeatedly  introduces $\epsilon$ edges to bypass  a pair of
%% consecutive edges labelled \bcar\acar\  or \bcdr\acdr.  The process is
%% continued  till a fixed  point is  reached.  When  the fixed  point is
%% reached,   the  resulting   NFA  contains   all   possible  reductions
%% corresponding to all the paths in the original NFA.  The proofs of the
%% termination and correctness of  th The algorithm repeatedly introduces
%% $\epsilon$  edges  to bypass  a  pair  of  consecutive edges  labelled
%% \bcar\acar\  or \bcdr\acdr.   The process  is continued  till  a fixed
%% point is reached.  When the  fixed point is reached, the resulting NFA
%% contains all possible reductions corresponding to all the paths in the
%% original NFA.  The proofs of the termination and correctness of th

%% \end{boxedminipage}
%%  \caption{Algorithm for transforming an NFA to  accept forward
%%    paths only.}\label{algo:simplify-NFA} \figrule
%% \end{figure*}

%% The algorithm repeatedly introduces  $\epsilon$ edges to bypass a pair
%% of consecutive edges labelled  \bcar\acar\ or \bcdr\acdr.  The process
%% is continued till  a fixed point is reached.  When  the fixed point is
%% reached,   the  resulting   NFA  contains   all   possible  reductions
%% corresponding to all the paths in the original NFA.  The proofs of the
%% termination and correctness  of the algorithm are given  in an earlier
%% paper~\cite{karkare07liveness}.

%%\input{liveness-dfa.tex}
\comment{change according  to example} We illustrate  the algorithm in
Figure~\ref{algo:simplify-NFA}  by   constructing  the  automaton  for
\var{\Lanv{9}{y}}.   Figure~\ref{fig:automata-construction}  (c) shows
the  automaton  for  \var{\Lanv{9}{y}}  constructed by  composing  the
automata   for    \var{\Uf{\append}{1}},   \var{\Df{\append}{1}}   and
$(\epsilon \mid \acdr \mid \acdr\acar\sigma_{\mathit{\!all}})$.  After
$\epsilon$ removal we get (d). We add an $\epsilon$ edge from $q_2$ to
$q_3$ bypassing the  \bcdr\acdr\ pair from $q_2$ to  $q_2$ and then to
$q_3$. This  is shown in (e).   The $\epsilon$ edge is  removed in (f)
and  a second  $\epsilon$  is  added to  the  automaton bypassing  the
\bcar\acar\ pair  from $q_1$ to $q_2$.  Removing  this $\epsilon$ edge
gives  the automaton  shown  in (g).   Restricting  this automaton  to
forward edges  only, we  get the final  automaton shown in  (h).  This
automaton  recognizes  $\acdr^{*}\!\mid\acdr^{*}\acar  \sigma_{\mathit
  \!all}$, showing  that the entire list \py,  including its elements,
is live at $\pi_9$. Also note  that the language accepted by the final
automaton satisfies the prefix-closed property.}
\begin{figure}[t!]
\hspace*{1cm}{\epsfig{file=example-automata.eps, height=2.75cm}}
$\caption{(a) The grammar rules for $\Df{\length}{1'}$ converted into
$  an automaton, and simplified in   (b). (c) and (d) show the automata
$  for   \Lanv{9}{\pa}.}
\figrule
\end{figure}

>>>>

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
%\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage [plain,noend,noline,boxed]{algorithm2e}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
%\usepackage{program}
%\usepackage{algorithmic}
\usepackage[boxed]{algorithm2e}

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\cred}[1]{{\color{red}{#1}}}
\newcommand{\blankout}[1]{}

%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } 
\newcommand{\stk}{\mbox{S}}   
\newcommand{\ID}{\mbox{$\mathbf{ id}$}}
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\begin{document}
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be
%over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data
%(0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Liveness-Based Garbage Collection for Lazy Languages}

% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.
\cmt{{
\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors
section.
%
\author{
% 1st. author
\alignauthor K. Prasanna Kumar \\
       \affaddr{IIT Bombay,}\\
       \affaddr{Mumbai 400076, India}
       \email{prasanna@cse.iitb.ac.in}
% 2nd. author
\alignauthor Amey Karkare\\
       \affaddr{IIT Kanpur}\\
       \affaddr{Kanpur 208016, India}
       \email{karkare@cse.iitk.ac.in}
% 3rd. author
\alignauthor Amitabha Sanyal \\
       \affaddr{IIT Bombay}\\
       \affaddr{Mumbai 400076, India}\\
       \email{as@cse.iitb.ac.in}
}
}}
\authorinfo{Double Blind Review}{}{}
\maketitle

@@

While  the runtime  system  of most  functional  languages includes  a
garbage  collector  (GC)  to  efficiently  reclaim  memory,  empirical
studies on Scheme~\cite{karkare06effectiveness} and, more importantly,
on Haskell~\cite{rojemo96lag}
%and Java~\cite{shaham02estimating}
programs have shown that  GCs leave uncollected a large
number of  memory objects that are  reachable but not  live (here {\em
  live} means the  object can potentially be used by  the program at a
later stage).  This results in unnecessary memory retention.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%% MOTIVATING EXAMPLE
%% \newcommand{\nilfigure}
%% {\scalebox{0.75}{
%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
%% \begin{pspicture}(0,0)(1,1)
%% %\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,
%% subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
%% \putnode{start}{origin}{0}{0}{}
%% \putnode{stop}{origin}{10}{10}{}
%% \ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
%% \end{pspicture}
%% }}

\begin{figure}[t!]
  %%%%{\color{Myblue}
  \psset{unit=1mm}
  \begin{pspicture}(80,50)(0,-50)
    %\psframe(80,50)(0,-50)
    %  \begin{center}
    \begin{tabular}{@{}c@{}}
      \begin{boxedminipage}{.42\textwidth}
        {\sf
	  \renewcommand{\arraystretch}{1}{
	    \begin{uprogram}
              \UFL\
              \UNL{0} (\DEFINE\ (\length\  \pl)
	      \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
              (\PRIM\ 1\ (\CAR\  \pl))))
              \UNL{0}
	      \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
	      \UNL{1}  (\SIF~(\NULLQ \ \lista)
	      \listb
	      \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\  \lista) (\append\
              (\CDR\  \lista)
              \listb))))
              \UNL{0}
              \UNL{0} (\LET\ \px\
              $\leftarrow$\ (\CONS\ $5$
              %          \UNL{1}\hspace*{1.5cm}
              (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
	      \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
	      \UNL{2}
              (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
              \UNL{3} (\SIF~(\NULLQ~(\CAR~(\CDR~\pz)))~$0$~$\pi$:\
              (\length\ \pz))))))
	    \end{uprogram}
        }}
      \end{boxedminipage}
      \\
      (a) Example program. \\ \\
      \begin{minipage}{.45\textwidth}
        %\hspace*{1cm}
                {\includegraphics[width=1\textwidth]{motiv-example}}
      \end{minipage}
      \\ \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
          (b) Memory graph at $\pi$. Thick edges denote live
          links. Traversal \\{\white (b)} stops at edges
          marked $\times$ during garbage collection.
      \end{tabular}}
    \end{tabular}
  \end{pspicture}
 \vspace*{-3ex}
  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
\end{figure}

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The analysis  has two  parts.  First,  each  function is
summarized in a context-sensitive manner as a {\em demand transformer}
that transforms a {\em symbolic} demand  on its body to demands on its
arguments.  This summary  is then used to step  through function calls
during analysis.   Second, the concrete  demand on a function  body is
obtained through a 0-CFA-like conservative approximation that combines
the  demands on  all the  calls to  the function.   The result  of the
analysis  is the annotation  of each  program point  with finite-state
automata capturing  the liveness of variables at  the point. Depending
on the program point where  garbage collection has been triggered, the
GC consults  a set of automata  to curtail reachability
during marking.  This results in  an increase in the garbage reclaimed
and consequently in fewer garbage collections.

@@

The edges shown by thick arrows  are those which are live at $\pi$.  A
cell is marked  and preserved during garbage collection  only if it is
reachable from the  root set through a path of  live edges.  All other
cells can be  reclaimed.  Thus if a garbage  collection takes place at
$\pi$  with  the  heap  shown  in  Figure~\ref{fig:mot-example}(b),  a
liveness-based collector (LGC) will  collect only the cells referenced
by  $\pz$, $(\CDR~  \pz)$  and (the  cells  constituting) the  closure
referenced by $(\CDR~(\CDR~\pz))$ need to be marked and preserved.  In
contrast,  a reachability-based collector  (RGC) would  have preserved
all cells.

@@

The      $\mathit{ref}$     rule      for     \CONS,      shown     in
Figure~\ref{fig:live-judge},  requires   us  to  remove   the  leading
\acar\ and \acdr\  from the access paths in  $\sigma$.  Similarly, the
rules for  \CAR, \CDR, \PRIM, \NULLQ,  and \SIF\ require  us to return
$\emptyset$, if  $\sigma$ itself is $\emptyset$ and
$\lbrace\epsilon\rbrace$ otherwise.  To  realize these rules
$\sigma$ needs to be known. This creates difficulties since we want to
solve the equations arising out of liveness symbolically.

The  solution  is  to   also  treat  the  operations  mentioned  above
symbolically.   We  introduce three  new  symbols.  These symbols  are
defined as a relation  $\hookrightarrow$ between sets of access paths.
First, \bcar\ and \bcdr\ are defined as:

\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \acdr\sigma_2'
\end{align*}
Thus \bcar, for example, removes the leading \acar\ from $\sigma_2$.
Similarly, the symbol \clazy\ reduces the string following it to an
$\epsilon$. \clazy\  is defined by addition to
$\hookrightarrow$ the rule:
\begin{align*}
  &\sigma_1\clazy\sigma_2  \hookrightarrow
  \sigma_1 \mbox{ if } \sigma_2 \subseteq  \{\acar,\acdr\}^\ast
\end{align*}
%% \begin{align*}
%%  \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
%%  \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
%% \end{align*}
%% and \clazy\ as:
%% \begin{align*}
%%  \clazy\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
%%                        & \{\epsilon\} ~\mbox{otherwise}
%% \end{align*}
%% Note that \clazy\ is idempotent.
We can  now rewrite the \CONS\  and the \CAR\  rules of $\mathit{ref}$
as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\clazy\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,\mathit{ep} \mapsto  x.\clazy \sigma]
\end{align*}
The rules for  \CDR, \PRIM\ and \NULLQ\ are also  modified in a manner
similar to \CAR.  %% The definitions  of \bcar\, \bcdr\ and \clazy\ induce
%% the following relation $\hookrightarrow$ over sets of access paths:
%% \begin{align*}
%%   &\sigma_1\bcar\sigma_2  \hookrightarrow
%%   \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
%% &\sigma_1\bcdr\sigma_2  \hookrightarrow
%%   \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
%%   \acdr\sigma_2',~\mbox{and}\\
%% &\sigma_1\clazy \sigma_2 \hookrightarrow \sigma_1, \mbox{ if }
%% \sigma_2 \ne
%%   \emptyset
%% \end{align*}
The reflexive  transitive closure of  $\hookrightarrow$ will
be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
following proposition relates  the original and the modified
liveness rules.

@@

Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the  grammar for  \var{\Df{\length}{1'}}.  The  NFA is
simplified by replacing the edges labeled \clazy\ by $\epsilon$. While
checking for liveness during GC, a forward access-path traversing such
an edge  will be accepted  only if it  can reach a final  state.  This
effectively implements  the $\hookrightarrow$ simplification  rule for
\clazy for forward paths. The resulting automaton, made deterministic,
is shown in Figure~\ref{fig:example-automata}(b). This expectedly says
that  given  a demand  $\epsilon$  on  \length,  the liveness  of  its
argument             is             $\acdr^{*}$.            Similarly,
Figure~\ref{fig:example-automata}(c)    shows   the    automaton   for
\var{\Lanv{9}{\pa}}. The simplified automaton, shown in
Figure~\ref{fig:example-automata}(d) does not accept any forward
paths. This reflects the lazy nature of our language. \length\  does not
evaluate its elements and thus \pa\ is dead at $\pi_9$.

\begin{figure}[t!]
\hspace*{1cm}{\epsfig{file=example-automata.eps, height=2.75cm}}
\caption{(a) The grammar rules for $\langle \Df{\length}{1'}\rangle$
converted into
  an automaton, and simplified in   (b). (c) and (d) show the automata
  for   \Lanv{9}{\pa}.}\label{fig:example-automata}
\figrule
\end{figure}
